---
title: "DSC: Compare fash, ftest and limma on simulated Gaussian data"
output: html_document
---
**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```

Simulate data from the Gaussian random effect model 
\[Y_{gi}=\mu_{gi}+\beta_{g,c(i)}+e_{gi}\]
where $\beta_{g,c}~N(0,\sigma_g^2)$, $e_{gi}~N(0,s_g^2)$.

Then we have F-statistics $F_g=\frac{MST_g}{MSE_g}=(1+N\sigma_g^2/s_g^2)\cdot F(df_1,df_2)$. Suppose $\alpha:=(1+N\sigma_g^2/s_g^2)$ and $\log(\alpha)$ comes from an non-negative unimodal prior (null: $\sigma=0 \Leftrightarrow \log(\alpha)=0$, alternative: non-negative unimodal prior with mode at 0).

Prior scenarios: spiky, near_normal, flat_top, big-normal, "bimodal" (actually not bimodal, unimodal at non-zero). 

Plot DSC results:
```{r,echo=FALSE,message=FALSE,out.width='\\textwidth',fig.height=10,fig.width=10}
library(ggplot2)
library(tidyr)
load("../code/dsc-fash/res.Rdata")
res = separate(res,scenario,c("scenario","nsamp","ngroup"),",")
res$nsamp = factor(res$nsamp, levels=c("nsamp=2","nsamp=10","nsamp=50"))
res$ngroup = factor(res$ngroup, levels=c("ngroup=3","ngroup=5","ngroup=10"))
res$scenario = factor(res$scenario, levels=c("spiky","near_normal","flat_top","big-normal"))

ggplot(res, aes(pi0,pi0.est,colour=method))+
  facet_grid(ngroup+nsamp~scenario) + geom_point(shape=1) +xlim(0,1) +ylim(0,1) + 
  xlab("true pi0") +ylab("estimated pi0") +geom_abline(slope=1,intercept=0,color=1)

ggplot(res, aes(pi0,FDP_005,colour=method)) + 
  facet_grid(ngroup+nsamp~scenario) + geom_point(shape=1) +xlim(0,1) +ylim(0,0.2) + 
  xlab("true pi0") +ylab("false discovery proportion when q=0.05") +geom_abline(slope=0,intercept=0.05,color=1)

ggplot(res, aes(pi0,DP_005,colour=method)) + 
  facet_grid(ngroup+nsamp~scenario) + geom_point(shape=1) +xlim(0,1) +ylim(0,1) + 
  xlab("true pi0") +ylab("discovery proportion when q=0.05") +geom_abline(slope=-1,intercept=1,color=1)

ggplot(res, aes(pi0,rmse.pve,colour=method)) + 
  facet_grid(ngroup+nsamp~scenario) + geom_point(shape=1) +xlim(0,1)+
  xlab("true pi0") +ylab("rmse.pve")

#####
library(dplyr)
newres = res
newres = newres[newres$method=='ftest',]
newres = newres[c(2:5,12)]
names(newres)[5] = c('rmse.pve_baseline')
test = left_join(res,newres,by=c('seed','scenario','ngroup','nsamp'))
test$rmse.pve_rel = test$rmse.pve/test$rmse.pve_baseline # RRMSE_prec

ggplot(test, aes(pi0,rmse.pve_rel,colour=method)) + 
  facet_grid(ngroup+nsamp~scenario) + geom_point(shape=1) +xlim(0,1)+
  xlab("true pi0") +ylab("relative rmse.pve")

```

Check an example where fash hardly improves PVE estimation accuracy compared to the baseline F-test:
```{r}
test[test[,1]=="big-normal,nsamp=2,ngroup=3"&test$method=="fash"&test$seed==15,c(1:2,7,19)]
```

Check fash's fitted prior (true prior: big-normal). 
```{r}
# Density of normal mixture
normalmix_dens = function(distn,xgrid){
  k = length(distn$pi)
  dens = colSums(dnorm(outer(rep(1,k),xgrid),distn$mu,distn$sd)*distn$pi)
  return(dens)
}

# Density of uniform mixture
unimix_dens = function(distn,xgrid){
  k = length(distn$pi)
  dens = colSums(dunif(outer(rep(1,k),xgrid),distn$a,distn$b)*distn$pi)
  return(dens)
}


meta=readRDS("../code/dsc-fash/dsc-fash-files/meta/big-normal,nsamp=2,ngroup=3/default_meta/meta.15.rds")
out.fash=readRDS("../code/dsc-fash/dsc-fash-files/output/big-normal,nsamp=2,ngroup=3/fash/fash_output/output.15.rds")

x = seq(-10,10,by=0.05)
fitted.alt = list(pi=out.fash$fit$fitted.g$pi[-1],a=out.fash$fit$fitted.g$a[-1],
                  b=out.fash$fit$fitted.g$b[-1])
plot(x[x>0],(normalmix_dens(meta$args$logaargs,x)[x>0])*2,type="l",ylim=c(0,0.5),xlim=c(0,10),
     ylab="prior density",xlab="x")
lines(x[x>0],unimix_dens(fitted.alt,x[x>0]),col=2)
legend("topright",lty=1,col=c(1,2),legend=c("true","fitted"))

```


## Session information

```{r info}
sessionInfo()
```
