---
title: "An example of using fash to reduce estimation MSE on simulated F-stats"
output: html_document
---
**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


We simulate F-statistics from the following model:
\begin{align}
\hat{F}_g &= \alpha_g \times F(df_1, df_2), \\
\log(\alpha_g) &\sim g(\cdot),
\end{align}
where $g(\cdot)$ is an non-negative unimodal prior for $\log(\alpha_g)$ with mode at 0: $g=\pi_0 \delta_0 + \pi_1 U(0, a)$.

We are interested in estimating parameters $\alpha_g$. In the linear regression model we have $\alpha_g=1+C\frac{\sigma_g^2}{s_g^2}$ where $\sigma^2$ is the variance of treatment effects and $s^2$ is the variance of errors. So $\alpha_g \geq 1$ ($\alpha_g=1$ iff $\sigma^2=0$) and that is why we let $g$ be non-negative.

```{r}
source('/Volumes/PERSONAL/MS/fash/fash.R')

library(qvalue)

# N: # of genes
# Nnull: # of nulls
# altsd: alternative prior component is U(-a,a) or U(0,a)
# oneside: flag, whether prior is positive (U(0,a)) or can be negative (U(-a,a))
sim_f = function(N,Nnull,df1,df2,altsd,oneside=TRUE){
  null = c(rep(1,Nnull),rep(0,N-Nnull))
  logalpha = rep(0,N)
  if (Nnull<N){
    if (oneside==TRUE){
      logalpha[(Nnull+1):N] = runif(N-Nnull,0,altsd)
    }else{
      logalpha[(Nnull+1):N] = runif(N-Nnull,-altsd,altsd)
    } 
  } 
  
  alpha = exp(logalpha)
  #Fhat = alpha*rf(N,df1=df1,df2=df2)
  logfhat = logalpha+log(rf(N,df1,df2))
  Fhat = exp(logfhat)
  return(list(Fhat=Fhat, logfhat=logfhat,
              logalpha=logalpha, null=null,
              df1=df1, df2=df2))
}
```

## A moderate df scenario (df1=10, df2=10)
```{r}
set.seed(999)

# oneside
data = sim_f(N=10000,Nnull=8000,df1=10,df2=10,altsd=4,oneside=TRUE)
test = fash(data$Fhat, df1=data$df1, df2=data$df2, 
                    oneside=TRUE, mixsd=4)
test$fitted.g

Fpvalue = 1-pf(data$Fhat,data$df1,data$df2)
qval.f = qvalue(Fpvalue)$qval
```

Compare the MSE's: fash gives much lower MSE (in estimating "\log(\alpha_g)").
```{r}
# MSE of F-test
mean((data$logalpha-log(data$Fhat))^2)

# MSE of fash
mean((data$logalpha-test$PosteriorMean.logf)^2)
```

Compare the FDR's of fash and raw F-test:

* Ranking of genes: almost same.
* F-test sometimes overestimates FDR, while fash can precisely estimate FDR.

```{r, echo=FALSE}
ng = 2000
plot((cumsum(data$null[order(test$qvalue)])/(1:length(test$qvalue)))[10:ng],
     type="l",xlab="# of top ranked genes",ylab="FDR",main="False discovery proportion (one-sided test)")
lines((cumsum(data$null[order(qval.f)])/(1:length(qval.f)))[10:ng],col=2)
legend("topleft",lty=1,col=c(1,2),legend=c("fash","F-test"))

plot((cumsum(data$null[order(test$qvalue)])/(1:length(test$qvalue))),
     sort(test$qvalue), xlab="true FDR", ylab="estimated FDR",main="fash")
abline(0,1)

plot((cumsum(data$null[order(qval.f)])/(1:length(qval.f))),
     sort(qval.f), xlab="true FDR", ylab="estimated FDR",main="F-test")
abline(0,1)
```

## A unbalanced df scenario (df1=5, df2=30)
```{r}
# oneside
data = sim_f(N=10000,Nnull=8000,df1=5,df2=30,altsd=4,oneside=TRUE)
test = fash(data$Fhat, df1=data$df1, df2=data$df2, 
                    oneside=TRUE, mixsd=4)
test$fitted.g

Fpvalue = 1-pf(data$Fhat,data$df1,data$df2)
qval.f = qvalue(Fpvalue)$qval
```

Compare the MSE's: fash gives lower MSE (in estimating "$\log(\alpha_g)$").
```{r}
# MSE of F-test
mean((data$logalpha-log(data$Fhat))^2)

# MSE of fash
mean((data$logalpha-test$PosteriorMean.logf)^2)
```

Compare the FDR's of fash and raw F-test: F-test sometimes slightly overestimates FDR, but fash doesn't.

```{r, echo=FALSE}
ng = 2000
plot((cumsum(data$null[order(test$qvalue)])/(1:length(test$qvalue)))[10:ng],
     type="l",xlab="# of top ranked genes",ylab="FDR",main="False discovery proportion (one-sided test)")
lines((cumsum(data$null[order(qval.f)])/(1:length(qval.f)))[10:ng],col=2)
legend("topleft",lty=1,col=c(1,2),legend=c("fash","F-test"))

plot((cumsum(data$null[order(test$qvalue)])/(1:length(test$qvalue))),
     sort(test$qvalue), xlab="true FDR", ylab="estimated FDR",main="fash")
abline(0,1)

plot((cumsum(data$null[order(qval.f)])/(1:length(qval.f))),
     sort(qval.f), xlab="true FDR", ylab="estimated FDR",main="F-test")
abline(0,1)
```

## A large df scenario (df1=50, df2=50)
```{r}
# oneside
data = sim_f(N=10000,Nnull=8000,df1=50,df2=50,altsd=4,oneside=TRUE)
test = fash(data$Fhat, df1=data$df1, df2=data$df2, 
                    oneside=TRUE, mixsd=4)
test$fitted.g

Fpvalue = 1-pf(data$Fhat,data$df1,data$df2)
qval.f = qvalue(Fpvalue)$qval
```

Compare the MSE's: fash gives lower MSE (in estimating "$\log(\alpha_g)$").
```{r}
# MSE of F-test
mean((data$logalpha-log(data$Fhat))^2)

# MSE of fash
mean((data$logalpha-test$PosteriorMean.logf)^2)
```

Compare the FDR's of fash and raw F-test: almost same (which makes sense since the sample size is already quite large).

```{r, echo=FALSE}
ng = 2000
plot((cumsum(data$null[order(test$qvalue)])/(1:length(test$qvalue)))[10:ng],
     type="l",xlab="# of top ranked genes",ylab="FDR",main="False discovery proportion (one-sided test)")
lines((cumsum(data$null[order(qval.f)])/(1:length(qval.f)))[10:ng],col=2)
legend("topleft",lty=1,col=c(1,2),legend=c("fash","F-test"))

plot((cumsum(data$null[order(test$qvalue)])/(1:length(test$qvalue))),
     sort(test$qvalue), xlab="true FDR", ylab="estimated FDR",main="fash")
abline(0,1)

plot((cumsum(data$null[order(qval.f)])/(1:length(qval.f))),
     sort(qval.f), xlab="true FDR", ylab="estimated FDR",main="F-test")
abline(0,1)
```


## Session information

```{r info}
sessionInfo()
```
