---
title: "A simulation example to compare variance decomposition by GLMM vs voom+ANOVA"
author: "Mengyin Lu"
date: 2016-02-10
---
**Last updated:** `r Sys.Date()`

**Code version:** `r system("git log -1 --format='%H'", intern = TRUE)`

```{r chunk-options, include=FALSE}
source("chunk-options.R")
```


We simulate the count matrix from the following GLMM model
\begin{align}
Y_{gi} &\sim Pois(\lambda_{gi}), \\
\log(\lambda_{gi}) &= \mu_g+\beta_{g,c(i)}+e_{gi}, \\
e_{gi}\sim N(0,\sigma_{ge}^2), &\quad \beta_{g,c(i)}\sim N(0,\sigma_{gc}^2).
\end{align}
where $\beta$ is the condition effect, $e$ is the random error (or sample effect). 

We define $r_g^{(GLMM)} := \frac{\sigma_{ge}^2}{\sigma_{gc}^2}$, which is the ratio of condition variation and sample variation. We are interested in estimating $r_g^{(GLMM)}$, or at least ranking the genes by this ratio. 

```{r}
set.seed(10000)

library(edgeR)
library(limma)

# Ngene: # of genes
# Nsamp: # of samples for each group
# K: number of groups
# logrmean, logrsd: generate true log(r)'s from N(logrmean, logrsd)
# mumean, musd: generate true mu's from N(logrmean, logrsd)
simcounts = function(Ngene, Nsamp, K, logrmean, logrsd, mumean, musd){
  sigma.e = rep(0.2,Ngene)
  logratio = rnorm(Ngene, mean=logrmean, sd=logrsd) #log(sigma.c^2/sigma.e^2)
  sigma.c = exp((logratio+2*log(sigma.e))/2)
  
  mu = rnorm(Ngene, mumean, musd)
  beta = matrix(rep(rnorm(Ngene*K, 0, rep(sigma.c^2,each=K)),each=Nsamp), nrow=Ngene, byrow=TRUE)
  e = matrix(rnorm(Ngene*K*Nsamp, 0, rep(sigma.e^2,each=K*Nsamp)), nrow=Ngene, byrow=TRUE)
  
  lambda = exp(mu+beta+e)
  
  counts = matrix(rpois(Ngene*K*Nsamp, lambda), nrow=Ngene)
  condition = factor(rep(1:K,each=Nsamp))
  return(list(counts=counts, condition=condition, logratio=logratio))
}

# Voom transformation & variance decomposition (ANOVA)
varcomp_voom = function(counts, condition){
  dgecounts = DGEList(counts=counts,group=condition)
  dgecounts = calcNormFactors(dgecounts)
  design = model.matrix(~condition)
  v = voom(dgecounts,design,plot=FALSE)
  dgecounts = estimateDisp(dgecounts, design, robust=TRUE)
  avelogcpm = dgecounts$AveLogCPM
  
  combo = cbind(v$E,v$weights)
  ratios = apply(combo,1,voom_ratio,condition=condition)
  return(list(logratio = log(ratios), avelogcpm=avelogcpm))
}

# r.voom is defined by ANOVA's MST/MSE
voom_ratio = function(ynweights,condition){
  y = ynweights[1:(length(ynweights)/2)]
  weights = ynweights[(length(ynweights)/2+1):length(ynweights)]
  fit = lm(y~condition, weights=weights)
  ratio = sqrt(max(anova(fit)$F[1]-1,0)/length(unique(condition)))
  return(ratio)
}

# Compute the ratios (sigma_c/sigma_e) for each gene
# Model: y~Pois(lambda), lambda=mu+b_cond+e, where b_cond~N(0,sigma_c^2), e~N(0,sigma_e^2)
varcomp_glmm = function(counts, condition){
  ratios = apply(counts,1,glmm_ratio,condition=condition)
  return(list(logratio = log(ratios)))
}

library(lme4)
glmm_ratio = function(y,condition){
  obs = factor(1:length(y))
  fit = try(glmer(y~(1|condition)+(1|obs),family=poisson))
  
  # Ratio = sigma_c/sigma_e
  if (class(fit)=="try-error"){
    ratio = NA
  }else{
    ratio = sqrt(as.numeric(summary(fit)$varcor$condition)/
                   as.numeric(summary(fit)$varcor$obs))
  }
  
  return(ratio)
}
```

We simulate a count dataset, with 20 groups and 20 samples for each group (so the sample size is sufficient big).

We can estimate $r_g^{(GLMM)}$'s by GLMM (REML). Another way is to run ANOVA on voom transformed data, and use $\hat{r}_g^{(voom)}:=MST_g/MSE_g$ to estimate the ratio of variations explained by condition effect & sample variation. Note that ANOVA is doing variance decomposition on log-cpm instead of the hidden Poisson mean space, so $r_g^{(voom)}$ and $r_g^{(GLMM)}$ are different quantities so not directly comparable! But we still hope the two ratios are correlated and have similar rankings. 

We compare the true $r_g^{(GLMM)}$'s against their GLMM estimates and $\hat{r}_g^{(voom)}$:

Voom+ANOVA: $\hat{r}_g^{(voom)}$'s are highly correlated with the true $r_g^{(GLMM)}$'s! Correlation>0.9.
GLMM(REML): Has some outliers (very inaccurate estimates, especially for low-expressed genes). Correlation between estimates and true log-ratios is less than 0.7. Possible drawbacks of REML: REML is unstable in estimating variance components especially at some boundry conditions, and the computation time is much slower. 

So maybe voom+ANOVA is a better choice? In this example it's more stable, faster and accurate. 
```{r}
data = simcounts(Ngene=1000, Nsamp=20, K=20, logrmean=0, logrsd=1, mumean=6, musd=0.5)
voomr = varcomp_voom(data$counts, data$condition)
glmmr = varcomp_glmm(data$counts, data$condition)

plot(data$logratio,voomr$logratio,xlab="true log(r)",
     ylab="voom+ANOVA estimated log(r)")
abline(0,1)
sum(voomr$logratio==-Inf)
cor(data$logratio[abs(voomr$logratio)!=Inf],voomr$logratio[abs(voomr$logratio)!=Inf])

plot(data$logratio,glmmr$logratio,xlab="true log(r)",
     ylab="GLMM estimated log(r)")
abline(0,1)
sum(glmmr$logratio==-Inf)
cor(data$logratio[abs(glmmr$logratio)!=Inf],glmmr$logratio[abs(glmmr$logratio)!=Inf])
```

Try another example with small sample size: 4 groups and 5 samples for each group. 
```{r}
data2 = simcounts(Ngene=1000, Nsamp=5, K=4, logrmean=0, logrsd=1.5, mumean=6, musd=1)
voomr2 = varcomp_voom(data2$counts, data2$condition)
glmmr2 = varcomp_glmm(data2$counts, data2$condition)

plot(data2$logratio,voomr2$logratio,xlab="true log(r)",
     ylab="voom+ANOVA estimated log(r)")
abline(0,1)
sum(voomr2$logratio==-Inf,na.rm=TRUE)
cor(data2$logratio[abs(voomr2$logratio)!=Inf],voomr2$logratio[abs(voomr2$logratio)!=Inf])

plot(data2$logratio,glmmr2$logratio,xlab="true log(r)",
     ylab="GLMM estimated log(r)")
abline(0,1)
sum(glmmr2$logratio==-Inf,na.rm=TRUE)
cor(data2$logratio[abs(glmmr2$logratio)!=Inf & !is.na(glmmr2$logratio)],
    glmmr2$logratio[abs(glmmr2$logratio)!=Inf& !is.na(glmmr2$logratio)])

```


## Session information

```{r info}
sessionInfo()
```
